---
title: "Iceland Municipalities Cartogram by Population"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Day 13 - A New Tool

The new tool is R and here we make a cartogram of the municipalities based on the population in different years.

You can find the geojson of the municipalities here: <https://gis.natt.is/geoserver/IS_50V/ows?service=WFS&version=1.0.0&request=GetFeature&typeName=IS_50V%3Amork_sveitarf_flakar&maxFeatures=100000&outputFormat=application%2Fjson>

\
You find the population data here: <https://px.hagstofa.is/pxis/pxweb/is/Ibuar/Ibuar__mannfjoldi__2_byggdir__sveitarfelog/MAN02005.px>

## Install Packages

```{r}
# Check and install required packages
packages <- c("sf", "dplyr", "readxl", "cartogram", "ggplot2", "gganimate", "magick")

# Install missing packages
installed_packages <- installed.packages()
for (pkg in packages) {
  if (!pkg %in% installed_packages) {
    install.packages(pkg)
  }
}

# Load all packages
lapply(packages, library, character.only = TRUE)
```

you can also load the packages like this:

```{r}
# Load necessary libraries
library(sf)
library(readxl)
# For cartogram
library(cartogram)
library(ggplot2)
library(dplyr)
# For video part
library(gganimate)
library(magick)
library(knitr)
```

## Load the population data

We assume the data has 1998 to 2024 year data to select

```{r}
# Load population data from Excel file
population_data <- read_excel("MAN02005_20241112-134452.xlsx", sheet = 1)
population_data <- population_data %>%
  select(Municipality = `Municipality`, starts_with("19"), starts_with("20"))
```

## Load the municipality polygons

We filter only the columns we need.

```{r}
# Load Icelandic municipality boundaries from GeoJSON, selecting only the required columns
municipalities <- st_read("mork_sveitarf_flakar.json") %>%
  select(id, nrsveitarfelags, sveitarfelag, geometry)

colnames(municipalities)
st_crs(municipalities)

# Transform to EPSG:3057 (ISN93 / Lambert Conformal Conic for Iceland)
municipalities_projected <- st_transform(municipalities, crs = 3857)
```

## Merge the datasets: Give population to the municipality polygons

```{r}
# Assuming 'population_data' is already loaded and prepared
# Merge spatial data with population data
merged_data <- municipalities_projected %>%
  left_join(population_data, by = c("sveitarfelag" = "Municipality"))
```

## Generate the cartogram

```{r}
# Generate cartogram based on 2024 population
cartogram_data <- cartogram_cont(merged_data, weight = "2024")
```

## Plot the cartogram for 2024

```{r plot-cartogram, echo=TRUE, fig.width=10, fig.height=8}
# Plot the cartogram
ggplot(cartogram_data) +
  geom_sf(aes(fill = Population_2024)) +
  scale_fill_viridis_c(option = "magma") +
  theme_minimal() +
  labs(title = "Icelandic Municipalities by Population in 2024",
       fill = "Population")
```

## View the data table (if needed)

```{r}
View(cartogram_data)
```

## Let's make it for each year (like frames for a video)

```{r}
# Define years for which we want to create cartograms
years <- as.character(2019:2024)
cartogram_frames <- list()

for (year in years) {
  # Merge the spatial data with population data for the current year
  merged_data <- municipalities %>%
    left_join(population_data %>% select(Municipality, Population = !!sym(year)),
              by = c("sveitarfelag" = "Municipality"))
  
  # Project the data for cartogram creation
  merged_data_projected <- st_transform(merged_data, crs = 3857)
  
  # Create cartogram
  cartogram_data <- cartogram_cont(merged_data_projected, weight = "Population")
  
  # Create plot
  p <- ggplot(cartogram_data) +
    geom_sf(aes(fill = Population)) +
    scale_fill_viridis_c(option = "magma") +
    theme_minimal() +
    labs(title = paste("Icelandic Municipalities by Population -", year),
         fill = "Population")
  
  # Save plot as a frame in the list
  cartogram_frames[[year]] <- p
}

```

```{r}
View(cartogram_frames)
```

## Save and Load processed data

```{r}
# Save the list of cartogram frames to a file
saveRDS(cartogram_frames, file = "cartogram_frames.rds")
```

```{r}
# Load the saved RDS data
cartogram_frames <- readRDS("cartogram_frames.rds")
```

## Adapting the styling and make GIF

```{r}
# Directory to save PNG files
output_dir <- "frames/"
dir.create(output_dir, showWarnings = FALSE)

# Save each frame as a PNG file with a year label, adjusted color scale, white background, and no axis labels
for (i in seq_along(cartogram_frames)) {
  year <- names(cartogram_frames)[i]
  file_path <- file.path(output_dir, paste0("cartogram_", year, ".png"))
  
  # Modify the plot to add the year label, remove grid and axis labels, add white background, and adjust color scale
  p <- cartogram_frames[[year]] +
    annotate("text", x = Inf, y = Inf, label = year, color = "orange", 
             size = 6, hjust = 1.1, vjust = 1.1, fontface = "bold") +
    theme_minimal() +
    theme(
      panel.grid = element_blank(),          # Remove grid lines
      axis.title = element_blank(),          # Remove axis titles
      axis.text = element_blank(),           # Remove axis text
      axis.ticks = element_blank(),          # Remove axis ticks
      panel.background = element_rect(fill = "lightblue", color = NA)
    ) +
    scale_fill_viridis_c(option = "magma")  # Accent low values
  
  # Save each ggplot frame as a PNG file
  ggsave(file_path, plot = p, width = 10, height = 8, dpi = 150, bg = "white")
}

# Load the images from C:/frames into magick and combine them into an animation
file_paths <- list.files(output_dir, pattern = "cartogram_.*\\.png$", full.names = TRUE)
file_paths <- sort(file_paths)  # Ensure they are ordered by year

# Create the GIF animation
animation <- image_read(file_paths) %>%
  image_animate(fps = 2)

gif = "Day13-ANewTool.gif"

# Save the GIF
image_write(animation, gif)
```

## Show the GIF

```{r, echo=FALSE}
# Load and display the GIF
animation <- image_read(gif)

# Show the GIF inline
include_graphics(gif)
```
